---
title: "Exploring New Hampshire School Enrollment Data"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Exploring New Hampshire School Enrollment Data}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 8,
  fig.height = 5,
  warning = FALSE,
  message = FALSE
)
```

## Status: Under Construction

The NH DOE iPlatform requires browser-based access, which prevents automated
data downloads. This package's data pipeline is built but **no real enrollment
data is currently bundled**. Once real data from the NH DOE is available, this
vignette will be updated with data stories and visualizations.

See the package README for manual data access instructions.

## Package API

```{r load-packages}
library(nhschooldata)
```

The main functions are ready to use once data is available:

```{r api-demo, eval=FALSE}
# Check available years
get_available_years()

# Fetch enrollment for a single year
enr <- fetch_enr(2024)

# Fetch multiple years
enr_multi <- fetch_enr_multi(2020:2024)

# Import manually downloaded file from NH DOE iPlatform
df <- import_local_enrollment("~/Downloads/enrollment.xlsx", 2024, "district")
```

## Manual Data Access

Download data directly from the NH DOE:

- **District enrollment**: <https://my.doe.nh.gov/iPlatform/Report/DataReportsSubCategory?reportSubCategoryId=9>
- **School enrollment**: <https://my.doe.nh.gov/iPlatform/Report/DataReportsSubCategory?reportSubCategoryId=10>

Then import using `import_local_enrollment()`:

```{r import-demo, eval=FALSE}
df <- import_local_enrollment(
  "~/Downloads/district-fall-enrollment.xlsx",
  end_year = 2024,
  level = "district"
)

# Process and tidy the imported data
processed <- process_enr(list(district = df, school = NULL), 2024)
tidy <- tidy_enr(processed)
```

## Session Info

```{r session-info}
sessionInfo()
```
