---
title: "Exploring New Hampshire School Enrollment Data"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Exploring New Hampshire School Enrollment Data}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 8,
  fig.height = 5,
  warning = FALSE,
  message = FALSE,
  cache = TRUE
)
```

```{r load-packages}
library(nhschooldata)
library(dplyr)
library(ggplot2)
library(tidyr)
```

```{r fetch-data}
enr <- fetch_enr_multi(2012:2026, use_cache = TRUE)
```

## State enrollment trend

New Hampshire's public school enrollment has declined steadily over the past
14 years. The state lost over 30,000 students — a 16% drop — even as charter
schools expanded and PreK programs grew.

### 1. NH lost 30,000 students since 2012

State enrollment fell from 190,805 to 160,322 — a 16% decline driven by
demographic contraction in one of America's oldest-population states.

```{r state-trend}
state_trend <- enr |>
  filter(is_state, subgroup == "total_enrollment", grade_level == "TOTAL") |>
  select(end_year, n_students) |>
  arrange(end_year)

stopifnot(nrow(state_trend) > 0)

state_trend |>
  mutate(
    change = n_students - lag(n_students),
    cumulative_change = n_students - first(n_students)
  )
```

```{r state-trend-chart}
ggplot(state_trend, aes(x = end_year, y = n_students)) +
  geom_line(linewidth = 1.2, color = "#1f77b4") +
  geom_point(size = 2, color = "#1f77b4") +
  scale_y_continuous(labels = scales::comma, limits = c(150000, 200000)) +
  scale_x_continuous(breaks = seq(2012, 2026, 2)) +
  labs(
    title = "New Hampshire Public School Enrollment, 2012-2026",
    subtitle = "Total enrollment declined 16% (-30,483 students)",
    x = "School Year (ending)",
    y = "Total Students",
    caption = "Source: NH DOE iPlatform"
  ) +
  theme_minimal(base_size = 13)
```

### 2. COVID erased 8,259 students in a single year

The 2020-21 school year saw a 4.7% enrollment drop — by far the largest
single-year decline in the dataset. NH has not recovered.

```{r covid-impact}
covid <- enr |>
  filter(is_state, subgroup == "total_enrollment", grade_level == "TOTAL",
         end_year %in% 2019:2022) |>
  select(end_year, n_students) |>
  arrange(end_year) |>
  mutate(
    yoy_change = n_students - lag(n_students),
    yoy_pct = round((n_students / lag(n_students) - 1) * 100, 1)
  )

stopifnot(nrow(covid) == 4)
covid
```

```{r covid-chart}
yoy <- enr |>
  filter(is_state, subgroup == "total_enrollment", grade_level == "TOTAL") |>
  select(end_year, n_students) |>
  arrange(end_year) |>
  mutate(yoy_change = n_students - lag(n_students)) |>
  filter(!is.na(yoy_change))

ggplot(yoy, aes(x = end_year, y = yoy_change, fill = yoy_change < 0)) +
  geom_col() +
  scale_fill_manual(values = c("TRUE" = "#d62728", "FALSE" = "#2ca02c"), guide = "none") +
  scale_x_continuous(breaks = seq(2013, 2026, 1)) +
  scale_y_continuous(labels = scales::comma) +
  labs(
    title = "Year-over-Year Change in NH Enrollment",
    subtitle = "COVID-19 caused an unprecedented 8,259-student drop in 2021",
    x = "School Year (ending)",
    y = "Change from Previous Year",
    caption = "Source: NH DOE iPlatform"
  ) +
  theme_minimal(base_size = 13) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

### 3. Charter schools grew from 0.6% to 3.9% of enrollment

While total enrollment shrank, charter schools quintupled their share — from
9 charters with 1,090 students in 2012 to 35 charters with 6,242 in 2026.

```{r charter-growth}
charter_names <- enr |>
  filter(is_district, end_year == 2026) |>
  filter(grepl("Charter|Chartered", district_name, ignore.case = TRUE)) |>
  distinct(district_name) |>
  pull()

charter_trend <- enr |>
  filter(is_district, subgroup == "total_enrollment", grade_level == "TOTAL",
         district_name %in% charter_names) |>
  group_by(end_year) |>
  summarize(
    n_charter = sum(n_students, na.rm = TRUE),
    n_charters = n(),
    .groups = "drop"
  )

state_total <- enr |>
  filter(is_state, subgroup == "total_enrollment", grade_level == "TOTAL") |>
  select(end_year, state_total = n_students)

charter_pct <- charter_trend |>
  left_join(state_total, by = "end_year") |>
  mutate(pct = round(n_charter / state_total * 100, 1))

stopifnot(nrow(charter_pct) > 0)
charter_pct
```

```{r charter-chart}
ggplot(charter_pct, aes(x = end_year)) +
  geom_col(aes(y = n_charter), fill = "#ff7f0e", alpha = 0.8) +
  geom_text(aes(y = n_charter, label = paste0(pct, "%")),
            vjust = -0.5, size = 3.5) +
  scale_x_continuous(breaks = seq(2012, 2026, 2)) +
  scale_y_continuous(labels = scales::comma) +
  labs(
    title = "NH Charter School Enrollment, 2012-2026",
    subtitle = "From 9 charters (0.6%) to 35 charters (3.9%) of total enrollment",
    x = "School Year (ending)",
    y = "Charter School Students",
    caption = "Source: NH DOE iPlatform"
  ) +
  theme_minimal(base_size = 13)
```

## District deep dives

### 4. Manchester lost nearly a quarter of its students

New Hampshire's largest city saw enrollment drop from 15,536 to 11,712 — a
loss of 3,824 students (24.6%) while maintaining 20 schools.

```{r manchester-nashua}
big2 <- enr |>
  filter(is_district, subgroup == "total_enrollment", grade_level == "TOTAL",
         district_name %in% c("Manchester", "Nashua")) |>
  select(end_year, district_name, n_students) |>
  arrange(end_year, district_name)

stopifnot(nrow(big2) > 0)
big2 |>
  filter(end_year %in% c(2012, 2018, 2026)) |>
  pivot_wider(names_from = district_name, values_from = n_students)
```

```{r manchester-nashua-chart}
ggplot(big2, aes(x = end_year, y = n_students, color = district_name)) +
  geom_line(linewidth = 1.1) +
  geom_point(size = 2) +
  scale_y_continuous(labels = scales::comma) +
  scale_x_continuous(breaks = seq(2012, 2026, 2)) +
  scale_color_manual(values = c("Manchester" = "#1f77b4", "Nashua" = "#d62728")) +
  labs(
    title = "Manchester vs Nashua: NH's Two Largest Districts",
    subtitle = "Both declining, but Manchester lost 3,824 students (-24.6%)",
    x = "School Year (ending)",
    y = "Total Students",
    color = "District",
    caption = "Source: NH DOE iPlatform"
  ) +
  theme_minimal(base_size = 13)
```

### 5. Errol has 12 students — the state's tiniest district

New Hampshire has dozens of districts with fewer than 100 students. The
smallest, Errol, has just 12 — smaller than most college seminar classes.

```{r smallest-districts}
smallest <- enr |>
  filter(is_district, subgroup == "total_enrollment", grade_level == "TOTAL",
         end_year == 2026) |>
  select(district_name, sau, sau_name, n_students) |>
  arrange(n_students) |>
  head(10)

stopifnot(nrow(smallest) > 0)
smallest
```

```{r smallest-chart}
smallest_plot <- smallest |>
  mutate(district_name = forcats::fct_reorder(district_name, n_students))

ggplot(smallest_plot, aes(x = district_name, y = n_students)) +
  geom_col(fill = "#2ca02c") +
  geom_text(aes(label = n_students), hjust = -0.3, size = 3.5) +
  coord_flip() +
  labs(
    title = "NH's 10 Smallest School Districts (2025-26)",
    subtitle = "Errol has just 12 students enrolled",
    x = NULL,
    y = "Total Students",
    caption = "Source: NH DOE iPlatform"
  ) +
  theme_minimal(base_size = 13)
```

### 6. Virtual Learning Academy grew 756% — the biggest gainer

While most districts shrank, Virtual Learning Academy Charter School grew
from 63 students in 2012 to 539 in 2026 — a 756% increase.

```{r biggest-gainers}
changes <- enr |>
  filter(is_district, subgroup == "total_enrollment", grade_level == "TOTAL",
         end_year %in% c(2012, 2026)) |>
  select(end_year, district_name, n_students) |>
  pivot_wider(names_from = end_year, values_from = n_students,
              names_prefix = "y") |>
  filter(!is.na(y2012), !is.na(y2026)) |>
  mutate(
    change = y2026 - y2012,
    pct_change = round((y2026 / y2012 - 1) * 100, 1)
  )

stopifnot(nrow(changes) > 0)
cat("Top 5 gainers:\n")
changes |> arrange(desc(pct_change)) |> head(5)
```

```{r gainers-chart}
top_gainers <- changes |>
  arrange(desc(change)) |>
  head(8) |>
  mutate(district_name = forcats::fct_reorder(district_name, change))

ggplot(top_gainers, aes(x = district_name, y = change, fill = change > 0)) +
  geom_col() +
  geom_text(aes(label = paste0(ifelse(change > 0, "+", ""), change)),
            hjust = -0.1, size = 3.5) +
  coord_flip() +
  scale_fill_manual(values = c("TRUE" = "#2ca02c"), guide = "none") +
  labs(
    title = "Districts That Grew Despite Statewide Decline",
    subtitle = "2012 to 2026, absolute student change",
    x = NULL,
    y = "Change in Students",
    caption = "Source: NH DOE iPlatform"
  ) +
  theme_minimal(base_size = 13)
```

### 7. Top 5 losers account for 9,481 lost students

The five largest districts — Manchester, Nashua, Hudson, Concord, and
Londonderry — together lost 9,481 students, nearly a third of the statewide
decline.

```{r biggest-losers}
top_losers <- changes |>
  arrange(change) |>
  head(5)

stopifnot(nrow(top_losers) == 5)
top_losers
```

```{r losers-chart}
losers_trend <- enr |>
  filter(is_district, subgroup == "total_enrollment", grade_level == "TOTAL",
         district_name %in% top_losers$district_name) |>
  select(end_year, district_name, n_students)

ggplot(losers_trend, aes(x = end_year, y = n_students, color = district_name)) +
  geom_line(linewidth = 1) +
  geom_point(size = 1.5) +
  scale_y_continuous(labels = scales::comma) +
  scale_x_continuous(breaks = seq(2012, 2026, 2)) +
  labs(
    title = "NH's 5 Biggest Losers: Enrollment Since 2012",
    subtitle = "Combined loss of 9,481 students (31% of statewide decline)",
    x = "School Year (ending)",
    y = "Total Students",
    color = "District",
    caption = "Source: NH DOE iPlatform"
  ) +
  theme_minimal(base_size = 13)
```

## Grade-level patterns

### 8. Grade 12 consistently outnumbers Grade 1

Every year since 2012, more students graduate from 12th grade than enter 1st
grade — a demographic signature of sustained population decline.

```{r pipeline-ratio}
pipeline <- enr |>
  filter(is_campus, subgroup == "total_enrollment",
         grade_level %in% c("01", "12")) |>
  group_by(end_year, grade_level) |>
  summarize(n = sum(n_students, na.rm = TRUE), .groups = "drop") |>
  pivot_wider(names_from = grade_level, values_from = n) |>
  rename(grade_01 = `01`, grade_12 = `12`) |>
  mutate(ratio = round(grade_12 / grade_01, 2))

stopifnot(nrow(pipeline) > 0)
pipeline
```

```{r pipeline-chart}
pipeline_long <- pipeline |>
  pivot_longer(cols = c(grade_01, grade_12), names_to = "grade",
               values_to = "students") |>
  mutate(grade = ifelse(grade == "grade_01", "Grade 1", "Grade 12"))

ggplot(pipeline_long, aes(x = end_year, y = students, color = grade)) +
  geom_line(linewidth = 1.1) +
  geom_point(size = 2) +
  scale_y_continuous(labels = scales::comma) +
  scale_x_continuous(breaks = seq(2012, 2026, 2)) +
  scale_color_manual(values = c("Grade 1" = "#2ca02c", "Grade 12" = "#9467bd")) +
  labs(
    title = "Grade 1 vs Grade 12 Enrollment",
    subtitle = "More students exit than enter NH public schools each year",
    x = "School Year (ending)",
    y = "Students",
    color = NULL,
    caption = "Source: NH DOE iPlatform (school-level data)"
  ) +
  theme_minimal(base_size = 13)
```

### 9. PreK enrollment grew 39% while everything else shrank

PreK enrollment rose from 3,165 to 4,395 — a 39% increase — even as overall
enrollment fell 16%. The COVID crash in 2021 hit PreK hardest (37% drop) but
it fully recovered by 2023.

```{r prek-trend}
prek <- enr |>
  filter(is_state, subgroup == "total_enrollment", grade_level == "PK") |>
  select(end_year, n_students) |>
  arrange(end_year)

stopifnot(nrow(prek) > 0)
prek
```

```{r prek-chart}
ggplot(prek, aes(x = end_year, y = n_students)) +
  geom_line(linewidth = 1.2, color = "#ff7f0e") +
  geom_point(size = 2, color = "#ff7f0e") +
  scale_y_continuous(labels = scales::comma) +
  scale_x_continuous(breaks = seq(2012, 2026, 2)) +
  labs(
    title = "PreK Enrollment Bucked the Statewide Decline",
    subtitle = "Grew 39% (3,165 to 4,395) while total enrollment fell 16%",
    x = "School Year (ending)",
    y = "PreK Students",
    caption = "Source: NH DOE iPlatform"
  ) +
  theme_minimal(base_size = 13)
```

### 10. Kindergarten lost 1,177 students since 2012

Kindergarten enrollment dropped from 11,904 to 10,727 — a 10% decline that
signals continued enrollment losses ahead.

```{r kindergarten}
k_trend <- enr |>
  filter(is_campus, subgroup == "total_enrollment", grade_level == "K") |>
  group_by(end_year) |>
  summarize(n_students = sum(n_students, na.rm = TRUE), .groups = "drop")

stopifnot(nrow(k_trend) > 0)
k_trend
```

```{r kindergarten-chart}
ggplot(k_trend, aes(x = end_year, y = n_students)) +
  geom_line(linewidth = 1.2, color = "#17becf") +
  geom_point(size = 2, color = "#17becf") +
  scale_y_continuous(labels = scales::comma) +
  scale_x_continuous(breaks = seq(2012, 2026, 2)) +
  labs(
    title = "Kindergarten Enrollment, 2012-2026",
    subtitle = "Down 10% — each incoming class is smaller than the last",
    x = "School Year (ending)",
    y = "Kindergarten Students",
    caption = "Source: NH DOE iPlatform (school-level data)"
  ) +
  theme_minimal(base_size = 13)
```

## Structural patterns

### 11. 111 one-school districts vs Manchester's 20

Over half of NH's districts have just one school. Meanwhile Manchester
operates 20 — nearly as many as the bottom 40 districts combined.

```{r school-counts}
schools_per <- enr |>
  filter(is_campus, subgroup == "total_enrollment", grade_level == "TOTAL",
         end_year == 2026) |>
  group_by(district_name) |>
  summarize(n_schools = n(), .groups = "drop") |>
  arrange(desc(n_schools))

stopifnot(nrow(schools_per) > 0)
cat("Districts by school count:\n")
schools_per |>
  mutate(category = case_when(
    n_schools == 1 ~ "1 school",
    n_schools <= 3 ~ "2-3 schools",
    n_schools <= 10 ~ "4-10 schools",
    TRUE ~ "11+ schools"
  )) |>
  count(category)
```

```{r school-counts-chart}
schools_hist <- schools_per |>
  mutate(bucket = cut(n_schools, breaks = c(0, 1, 3, 5, 10, 20),
                      labels = c("1", "2-3", "4-5", "6-10", "11-20")))

ggplot(schools_hist, aes(x = bucket)) +
  geom_bar(fill = "#1f77b4") +
  geom_text(stat = "count", aes(label = after_stat(count)), vjust = -0.5) +
  labs(
    title = "How Many Schools Do NH Districts Operate? (2025-26)",
    subtitle = "111 districts have just one school; Manchester has 20",
    x = "Number of Schools",
    y = "Number of Districts",
    caption = "Source: NH DOE iPlatform"
  ) +
  theme_minimal(base_size = 13)
```

### 12. The SAU system: one administrator, multiple districts

New Hampshire's School Administrative Units (SAUs) are a unique feature.
Plymouth SAU #48 covers 8 districts, allowing tiny towns to share
administrative costs.

```{r sau-system}
sau_multi <- enr |>
  filter(is_district, subgroup == "total_enrollment", grade_level == "TOTAL",
         end_year == 2026) |>
  group_by(sau, sau_name) |>
  summarize(
    n_districts = n(),
    total_students = sum(n_students, na.rm = TRUE),
    districts = paste(district_name, collapse = ", "),
    .groups = "drop"
  ) |>
  arrange(desc(n_districts))

stopifnot(nrow(sau_multi) > 0)
sau_multi |> head(8)
```

```{r sau-chart}
sau_top <- sau_multi |>
  head(10) |>
  mutate(sau_label = paste0("SAU ", sau, " (", sau_name, ")")) |>
  mutate(sau_label = forcats::fct_reorder(sau_label, n_districts))

ggplot(sau_top, aes(x = sau_label, y = n_districts)) +
  geom_col(fill = "#9467bd") +
  geom_text(aes(label = paste0(n_districts, " districts\n(",
                                scales::comma(total_students), " students)")),
            hjust = -0.05, size = 3) +
  coord_flip() +
  labs(
    title = "SAUs with the Most Districts (2025-26)",
    subtitle = "Plymouth SAU #48 covers 8 independent districts",
    x = NULL,
    y = "Number of Districts",
    caption = "Source: NH DOE iPlatform"
  ) +
  theme_minimal(base_size = 13) +
  theme(plot.margin = margin(5.5, 40, 5.5, 5.5))
```

### 13. North Country lost 26% of its students

The remote northern districts (Coos County area) saw enrollment drop from
2,526 to 1,859 — a 26% decline, steeper than the statewide average.

```{r north-country}
north_country <- c("Berlin", "Gorham", "Milan", "Errol", "Pittsburg",
                    "Colebrook", "Stark", "Stratford", "Stewartstown",
                    "Northumberland", "Lancaster", "Whitefield", "Dalton")

nc_trend <- enr |>
  filter(is_district, subgroup == "total_enrollment", grade_level == "TOTAL",
         district_name %in% north_country) |>
  group_by(end_year) |>
  summarize(
    n_students = sum(n_students, na.rm = TRUE),
    n_districts = n(),
    .groups = "drop"
  )

stopifnot(nrow(nc_trend) > 0)
nc_trend
```

```{r north-country-chart}
ggplot(nc_trend, aes(x = end_year, y = n_students)) +
  geom_line(linewidth = 1.2, color = "#d62728") +
  geom_point(size = 2, color = "#d62728") +
  scale_y_continuous(labels = scales::comma) +
  scale_x_continuous(breaks = seq(2012, 2026, 2)) +
  labs(
    title = "North Country Enrollment: Faster Decline Than the State",
    subtitle = "Coos County area districts lost 26% of students (2,526 to 1,859)",
    x = "School Year (ending)",
    y = "Total Students",
    caption = "Source: NH DOE iPlatform"
  ) +
  theme_minimal(base_size = 13)
```

### 14. Southern NH's Boston corridor also losing students

Even districts near the Massachusetts border — traditionally NH's growth
engine — are declining. Salem, Windham, Londonderry, and Hudson together lost
3,432 students.

```{r southern-nh}
south_nh <- c("Salem", "Windham", "Londonderry", "Derry", "Hudson",
              "Pelham", "Hampstead", "Atkinson", "Plaistow", "Sandown")

snh_trend <- enr |>
  filter(is_district, subgroup == "total_enrollment", grade_level == "TOTAL",
         district_name %in% south_nh) |>
  group_by(end_year) |>
  summarize(n_students = sum(n_students, na.rm = TRUE), .groups = "drop")

stopifnot(nrow(snh_trend) > 0)
snh_trend |> filter(end_year %in% c(2012, 2020, 2026))
```

```{r southern-nh-chart}
# Compare Southern NH vs State (indexed to 2012 = 100)
state_idx <- state_trend |>
  mutate(index = round(n_students / first(n_students) * 100, 1),
         region = "Statewide")

snh_idx <- snh_trend |>
  mutate(index = round(n_students / first(n_students) * 100, 1),
         region = "Southern NH (Boston corridor)")

nc_idx <- nc_trend |>
  mutate(index = round(n_students / first(n_students) * 100, 1),
         region = "North Country")

combined <- bind_rows(
  state_idx |> select(end_year, index, region),
  snh_idx |> select(end_year, index, region),
  nc_idx |> select(end_year, index, region)
)

ggplot(combined, aes(x = end_year, y = index, color = region)) +
  geom_line(linewidth = 1.1) +
  geom_point(size = 2) +
  geom_hline(yintercept = 100, linetype = "dashed", alpha = 0.5) +
  scale_x_continuous(breaks = seq(2012, 2026, 2)) +
  labs(
    title = "Regional Enrollment Decline (Indexed to 2012 = 100)",
    subtitle = "North Country declining fastest, but even Southern NH is down",
    x = "School Year (ending)",
    y = "Index (2012 = 100)",
    color = NULL,
    caption = "Source: NH DOE iPlatform"
  ) +
  theme_minimal(base_size = 13) +
  theme(legend.position = "bottom")
```

### 15. The grade pyramid: where NH's students are

The 2025-26 grade distribution shows that upper grades have more students
than lower grades — confirming the downward demographic trend. Grade 9 has
the most students (13,156), while Grade 1 has the fewest K-12 (11,169).

```{r grade-pyramid}
grade_dist <- enr |>
  filter(is_campus, subgroup == "total_enrollment",
         end_year == 2026,
         !grade_level %in% c("TOTAL", "ELEM", "MIDDLE", "HIGH")) |>
  group_by(grade_level) |>
  summarize(n_students = sum(n_students, na.rm = TRUE), .groups = "drop") |>
  arrange(grade_level)

stopifnot(nrow(grade_dist) > 0)
grade_dist
```

```{r grade-pyramid-chart}
grade_order <- c("PK", "K", "01", "02", "03", "04", "05", "06",
                 "07", "08", "09", "10", "11", "12", "PG")
grade_labels <- c("PreK", "K", "1", "2", "3", "4", "5", "6",
                  "7", "8", "9", "10", "11", "12", "PG")

grade_plot <- grade_dist |>
  mutate(
    grade_level = factor(grade_level, levels = grade_order, labels = grade_labels)
  ) |>
  filter(!is.na(grade_level))

ggplot(grade_plot, aes(x = grade_level, y = n_students)) +
  geom_col(fill = "#1f77b4") +
  geom_text(aes(label = scales::comma(n_students)), vjust = -0.3, size = 3) +
  scale_y_continuous(labels = scales::comma) +
  labs(
    title = "NH Enrollment by Grade Level (2025-26)",
    subtitle = "Upper grades larger than lower — demographic contraction in real time",
    x = "Grade",
    y = "Students",
    caption = "Source: NH DOE iPlatform (school-level data)"
  ) +
  theme_minimal(base_size = 13)
```

## Session Info

```{r session-info}
sessionInfo()
```
