---
title: "Exploring New Hampshire School Enrollment Data"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Exploring New Hampshire School Enrollment Data}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 8,
  fig.height = 5,
  warning = FALSE,
  message = FALSE,
  cache = TRUE
)
```

## Data Access Note

The NH DOE iPlatform requires browser-based access. If automatic download fails:

1. Visit [my.doe.nh.gov/iPlatform](https://my.doe.nh.gov/iPlatform)
2. Download the enrollment reports
3. Use `import_local_enrollment()` to load the data

```{r load-packages}
library(nhschooldata)
library(dplyr)
library(tidyr)
library(ggplot2)
theme_set(theme_minimal(base_size = 14))
```

## 1. Explore statewide enrollment trends

Track how New Hampshire's public school enrollment has changed over time.

```{r statewide-data}
enr <- fetch_enr_multi(2016:2024, use_cache = TRUE)

statewide <- enr %>%
  filter(is_state, subgroup == "total_enrollment", grade_level == "TOTAL") %>%
  select(end_year, n_students)

statewide
```

```{r statewide-chart, fig.cap="New Hampshire statewide enrollment 2016-2024"}
ggplot(statewide, aes(x = end_year, y = n_students)) +
  geom_line(color = "#2E86AB", linewidth = 1.2) +
  geom_point(color = "#2E86AB", size = 3) +
  scale_y_continuous(labels = scales::comma, limits = c(0, NA)) +
  labs(
    title = "New Hampshire Public School Enrollment (2016-2024)",
    subtitle = "Tracking statewide trends",
    x = "Year",
    y = "Students"
  )
```

## 2. Find the largest districts

See which districts serve the most students in New Hampshire.

```{r top-districts-data}
enr_2024 <- fetch_enr(2024, use_cache = TRUE)

top_districts <- enr_2024 %>%
  filter(is_district, subgroup == "total_enrollment", grade_level == "TOTAL") %>%
  arrange(desc(n_students)) %>%
  head(10) %>%
  select(district_name, n_students)

top_districts
```

```{r top-districts-chart, fig.cap="Top 10 New Hampshire school districts by enrollment"}
top_districts %>%
  mutate(district_name = gsub(" School District", "", district_name)) %>%
  mutate(district_name = factor(district_name, levels = rev(district_name))) %>%
  ggplot(aes(x = n_students, y = district_name)) +
  geom_col(fill = "#2E86AB") +
  geom_text(aes(label = scales::comma(n_students)), hjust = -0.1, size = 3.5) +
  scale_x_continuous(labels = scales::comma, expand = expansion(mult = c(0, 0.15))) +
  labs(
    title = "Largest School Districts in New Hampshire (2024)",
    x = "Students",
    y = NULL
  )
```

## 3. Track demographic changes

Monitor how the demographic composition of New Hampshire schools is changing.

```{r demographics-data}
enr_demo <- fetch_enr_multi(c(2016, 2018, 2020, 2024), use_cache = TRUE)

demographics <- enr_demo %>%
  filter(is_state, grade_level == "TOTAL",
         subgroup %in% c("white", "hispanic", "asian", "black", "multiracial")) %>%
  group_by(end_year) %>%
  mutate(pct = n_students / sum(n_students) * 100) %>%
  ungroup() %>%
  select(end_year, subgroup, pct)

demographics_wide <- demographics %>%
  pivot_wider(names_from = subgroup, values_from = pct)

demographics_wide
```

```{r demographics-chart, fig.cap="Demographic shifts in New Hampshire schools"}
demographics %>%
  mutate(subgroup = factor(subgroup,
                           levels = c("white", "hispanic", "asian", "multiracial", "black"),
                           labels = c("White", "Hispanic", "Asian", "Multiracial", "Black"))) %>%
  ggplot(aes(x = end_year, y = pct, color = subgroup)) +
  geom_line(linewidth = 1.2) +
  geom_point(size = 3) +
  scale_color_manual(values = c("White" = "#2E86AB", "Hispanic" = "#E94F37",
                                "Asian" = "#F6AE2D", "Multiracial" = "#86BA90",
                                "Black" = "#A23B72")) +
  scale_y_continuous(labels = function(x) paste0(x, "%")) +
  labs(
    title = "Demographic Composition of NH Public Schools (2016-2024)",
    x = "Year",
    y = "Percent of Students",
    color = "Group"
  ) +
  theme(legend.position = "right")
```

## 4. Compare regional trends

New Hampshire's geography creates distinct regional patterns - compare coastal areas to the North Country.

```{r regional-data}
seacoast <- c("Portsmouth", "Dover", "Rochester", "Exeter", "Hampton")
north_country <- c("Berlin", "Colebrook", "Lancaster", "Littleton", "Gorham")

enr_regional <- fetch_enr_multi(c(2016, 2018, 2021, 2024), use_cache = TRUE)

regional <- enr_regional %>%
  filter(is_district, subgroup == "total_enrollment", grade_level == "TOTAL") %>%
  mutate(region = case_when(
    grepl(paste(seacoast, collapse = "|"), district_name) ~ "Seacoast",
    grepl(paste(north_country, collapse = "|"), district_name) ~ "North Country",
    TRUE ~ "Other"
  )) %>%
  filter(region %in% c("Seacoast", "North Country")) %>%
  group_by(end_year, region) %>%
  summarize(total = sum(n_students, na.rm = TRUE), .groups = "drop")

regional
```

```{r regional-chart, fig.cap="Regional enrollment trends in New Hampshire"}
ggplot(regional, aes(x = end_year, y = total, color = region)) +
  geom_line(linewidth = 1.2) +
  geom_point(size = 3) +
  scale_color_manual(values = c("Seacoast" = "#2E86AB", "North Country" = "#E94F37")) +
  scale_y_continuous(labels = scales::comma) +
  labs(
    title = "Regional Enrollment: Seacoast vs. North Country",
    x = "Year",
    y = "Students",
    color = "Region"
  ) +
  theme(legend.position = "right")
```

## 5. Analyze grade-level trends

Track how enrollment varies across grade levels to understand the student pipeline.

```{r grade-level-data}
enr_grades <- fetch_enr_multi(2016:2024, use_cache = TRUE)

grades <- enr_grades %>%
  filter(is_state, subgroup == "total_enrollment",
         grade_level %in% c("K", "01", "05", "09", "12")) %>%
  select(end_year, grade_level, n_students) %>%
  mutate(grade_label = case_when(
    grade_level == "K" ~ "Kindergarten",
    grade_level == "01" ~ "1st Grade",
    grade_level == "05" ~ "5th Grade",
    grade_level == "09" ~ "9th Grade",
    grade_level == "12" ~ "12th Grade"
  ))

grades
```

```{r grade-level-chart, fig.cap="Enrollment by grade level over time"}
grades %>%
  mutate(grade_label = factor(grade_label,
                              levels = c("Kindergarten", "1st Grade", "5th Grade",
                                         "9th Grade", "12th Grade"))) %>%
  ggplot(aes(x = end_year, y = n_students, color = grade_label)) +
  geom_line(linewidth = 1) +
  geom_point(size = 2) +
  scale_color_brewer(palette = "Set1") +
  scale_y_continuous(labels = scales::comma) +
  labs(
    title = "New Hampshire Enrollment by Grade Level (2016-2024)",
    x = "Year",
    y = "Students",
    color = "Grade"
  ) +
  theme(legend.position = "right")
```

## 6. Understand district size distribution

New Hampshire has many small districts that share administration through SAUs.

```{r district-size-data}
district_sizes <- enr_2024 %>%
  filter(is_district, subgroup == "total_enrollment", grade_level == "TOTAL") %>%
  mutate(size = case_when(
    n_students >= 3000 ~ "Large (3,000+)",
    n_students >= 1000 ~ "Medium (1,000-2,999)",
    n_students >= 300 ~ "Small (300-999)",
    TRUE ~ "Tiny (<300)"
  )) %>%
  count(size) %>%
  mutate(size = factor(size, levels = c("Tiny (<300)", "Small (300-999)",
                                         "Medium (1,000-2,999)", "Large (3,000+)")))

district_sizes
```

```{r district-size-chart, fig.cap="Distribution of district sizes in New Hampshire"}
ggplot(district_sizes, aes(x = size, y = n, fill = size)) +
  geom_col() +
  geom_text(aes(label = n), vjust = -0.5, size = 4) +
  scale_fill_manual(values = c("Tiny (<300)" = "#E94F37", "Small (300-999)" = "#F6AE2D",
                               "Medium (1,000-2,999)" = "#86BA90", "Large (3,000+)" = "#2E86AB")) +
  scale_y_continuous(expand = expansion(mult = c(0, 0.1))) +
  labs(
    title = "New Hampshire Districts by Size (2024)",
    x = "District Size",
    y = "Number of Districts"
  ) +
  theme(legend.position = "none")
```

## 7. Compare Manchester and Nashua

Track the state's two largest cities and their share of total enrollment.

```{r big-cities-data}
big_cities <- enr %>%
  filter(is_district, subgroup == "total_enrollment", grade_level == "TOTAL",
         grepl("Manchester|Nashua", district_name)) %>%
  group_by(end_year) %>%
  summarize(combined = sum(n_students, na.rm = TRUE), .groups = "drop")

state_totals <- enr %>%
  filter(is_state, subgroup == "total_enrollment", grade_level == "TOTAL") %>%
  select(end_year, state_total = n_students)

big_cities %>%
  left_join(state_totals, by = "end_year") %>%
  mutate(pct_of_state = round(combined / state_total * 100, 1))
```

## 8. Track kindergarten pipeline

Monitor kindergarten enrollment to predict future enrollment trends.

```{r kindergarten-data}
kindergarten <- enr %>%
  filter(is_state, subgroup == "total_enrollment", grade_level == "K") %>%
  select(end_year, k_students = n_students)

# Calculate year-over-year change
kindergarten %>%
  mutate(
    change = k_students - lag(k_students),
    pct_change = round(change / lag(k_students) * 100, 1)
  )
```

## 9. Find the smallest districts

Identify New Hampshire's tiny districts that rely on SAU administrative sharing.

```{r tiny-districts-data}
tiny_districts <- enr_2024 %>%
  filter(is_district, subgroup == "total_enrollment", grade_level == "TOTAL",
         n_students < 200) %>%
  arrange(n_students) %>%
  head(10) %>%
  select(district_name, n_students)

tiny_districts
```

## 10. Track high school enrollment

Follow high school grades (9-12) to understand graduation pipeline.

```{r high-school-data}
high_school <- enr %>%
  filter(is_state, subgroup == "total_enrollment",
         grade_level %in% c("09", "10", "11", "12")) %>%
  group_by(end_year) %>%
  summarize(hs_total = sum(n_students, na.rm = TRUE), .groups = "drop")

high_school
```

## 11. Compare elementary vs secondary

Track the balance between elementary (K-5) and secondary (6-12) enrollment.

```{r elem-sec-data}
elem_grades <- c("K", "01", "02", "03", "04", "05")
sec_grades <- c("06", "07", "08", "09", "10", "11", "12")

level_comparison <- enr %>%
  filter(is_state, subgroup == "total_enrollment") %>%
  mutate(level = case_when(
    grade_level %in% elem_grades ~ "Elementary",
    grade_level %in% sec_grades ~ "Secondary",
    TRUE ~ NA_character_
  )) %>%
  filter(!is.na(level)) %>%
  group_by(end_year, level) %>%
  summarize(students = sum(n_students, na.rm = TRUE), .groups = "drop")

level_comparison
```

## 12. Identify growing districts

Find districts that have grown over the analysis period.

```{r growth-data}
growth <- enr %>%
  filter(is_district, subgroup == "total_enrollment", grade_level == "TOTAL",
         end_year %in% c(2016, 2024)) %>%
  select(district_name, end_year, n_students) %>%
  pivot_wider(names_from = end_year, values_from = n_students,
                     names_prefix = "y")

# Calculate growth if data exists
if (nrow(growth) > 0 && "y2024" %in% names(growth) && "y2016" %in% names(growth)) {
  growth <- growth %>%
    mutate(
      change = y2024 - y2016,
      pct_change = round(change / y2016 * 100, 1)
    ) %>%
    arrange(desc(pct_change)) %>%
    head(10)
}

growth
```

## 13. Track middle grades

Monitor grades 6-8 enrollment trends.

```{r middle-grades-data}
middle_grades <- enr %>%
  filter(is_state, subgroup == "total_enrollment",
         grade_level %in% c("06", "07", "08")) %>%
  group_by(end_year) %>%
  summarize(middle_total = sum(n_students, na.rm = TRUE), .groups = "drop")

middle_grades
```

## 14. Compare school vs district counts

See how many schools operate within each district size category.

```{r school-counts-data}
school_counts <- enr_2024 %>%
  filter(is_campus, subgroup == "total_enrollment", grade_level == "TOTAL") %>%
  group_by(district_name) %>%
  summarize(n_schools = n(), .groups = "drop")

district_enrollment <- enr_2024 %>%
  filter(is_district, subgroup == "total_enrollment", grade_level == "TOTAL") %>%
  select(district_name, district_enrollment = n_students)

school_counts %>%
  left_join(district_enrollment, by = "district_name") %>%
  mutate(avg_school_size = round(district_enrollment / n_schools, 0)) %>%
  arrange(desc(n_schools)) %>%
  head(10)
```

## 15. Calculate year-over-year state change

Track annual enrollment changes at the state level.

```{r state-changes-data}
state_changes <- enr %>%
  filter(is_state, subgroup == "total_enrollment", grade_level == "TOTAL") %>%
  arrange(end_year) %>%
  mutate(
    change = n_students - lag(n_students),
    pct_change = round(change / lag(n_students) * 100, 2)
  ) %>%
  select(end_year, n_students, change, pct_change)

state_changes
```

## Summary

New Hampshire's public school enrollment data reveals:

- **Statewide trends**: Track enrollment changes across years
- **District comparisons**: Compare districts by size and location
- **Demographics**: Monitor changing student populations
- **Regional patterns**: Understand Seacoast vs. North Country differences
- **Grade-level analysis**: Follow the student pipeline
- **Growth analysis**: Identify growing and declining districts

## Session Info

```{r session-info}
sessionInfo()
```
